---
title: "Proteome analyses"
output: html_notebook
author: "Zandra Fagernaes"
---

This is a walkthrough of all the proteome analyses for the paper "A unified protocol for simultaneous extraction of DNA and proteins from archaeological dental calculus" (Fagernaes et al. 2020). The data is summarized in figure 8. Individual RUV001 in not included in any statistics, as stated in the paper.

```{r, include=FALSE}
library(janitor)
library(readxl)
library(vegan)
library(ggpubr)
library(broom)
library(cowplot)
library(tidyverse)
```


# Protein number and cellular location

## Prepare data

The input file is the 'samples view report' exported from a Scaffold file where GO annotations are shown. Note that Scaffold inserts a few lines above the header, and a line saying 'end of file' at the end, which I removed manually before starting the analyses.

```{r}
# Load data 
raw_data <- read.delim("<PATH_TO_FILE>") %>%
  janitor::clean_names() 
metadata <- read.delim("<PATH_TO_FILE>")

# Select only relevant columns
# Note that ending numbers for identified proteins changes if the file is changed!
data <- raw_data %>%
  select(c("identified_proteins_786_797", "accession_number", 29:42, 62:85)) 
colnames(data)[colnames(data)=="identified_proteins_786_797"] <- "identified_proteins"

# Remove decoy hits, collagens and keratins
data <- data %>%
  filter(!str_detect(accession_number, 'rr//|REV')) %>%
  filter(!str_detect(identified_proteins, "Collagen|Keratin|Serum albumin"))
```

## Modify GO-term columns

Scaffold annotates proteins with GO-terms of cellular location, but they are nested and at various levels in the hierarchy. The output has many columns of different GO terms, with futher GO terms defined in the columns. Therefore,  we will summarize them so that only five main groups will remain:

1. Intracellular/organelle (golgi, cytoskeleton, ER, endosome, intracellular organelle,     
    mitochondrion, nucleus, organelle part, cytoplasm and ribosome columns)
2. Membrane (membrane, plasma membrane and organelle membrane)
3. Extracellular region
4. Various (i.e. found in more than one of locations 1-3)
5. NA (Information not available)

Categories were defined by looking at ancestral terms for every column of GO terms at https://www.ebi.ac.uk/QuickGO/

```{r}
# Substitute GO terms in every column with the number 1
data[ ,3:16] <- lapply(data[ ,3:16], function (x) gsub(" ", "", x))
data[ ,3:16] <- lapply(data[ ,3:16], function (x) gsub("[[:graph:]]+", "1", x))
data[ ,3:16] <- lapply(data[ ,3:16], as.numeric)

# Combine GO categories into group 1-3

## 1. Intracellular/organelle
data$group_intra_organ <- data %>% 
  select("golgi_apparatus", "cytoskeleton", "endoplasmic_reticulum", "endosome", "intracellular_organelle", "mitochondrion", "nucleus", "organelle_part", "cytoplasm", "ribosome") %>%
  rowSums(na.rm=TRUE)

## 2. Membranes
data$group_membrane <- data %>%
  select("membrane", "plasma_membrane", "organelle_membrane") %>%
  rowSums(na.rm=TRUE)

## 3. Extracellular
data$group_extra <- data %>%
  select("extracellular_region") %>%
  rowSums(na.rm=TRUE)

# Select only relevant columns for downstream analyses
data_go <- data %>%
  select("identified_proteins", "accession_number", "group_intra_organ", "group_membrane", "group_extra", 17:40)

# Change count numbers in GO-groups to 0 or 1 (1 = anything larger than 0)
data_go[ ,3:5] <- lapply(data_go[ ,3:5], function (x) gsub("^[1-9][0-9]*", "1", x))
```

## Assign one GO group to each protein cluster

Now that our GO-terms are in a neat form, we need to assign just one GO-group to each protein cluster. We do this by creating a sum-column of the three groups. If it is 2-3, the cluster has more than one cellular location, and is classified into 4) Various. If it is 0, the cluster has no GO-location assigned, and gets classified into 5) NA. 

```{r}
# Create column for GO group, where groups 1-3 are summed
data_go[3:5] <- lapply(data_go[3:5], as.numeric)
data_go$group_var_na <- data_go %>%
  select("group_intra_organ", "group_membrane", "group_extra") %>%
  rowSums()

# Assign groups 1-3
data_go$GOgroup[data_go$group_intra_organ=="1"] <- "intra_organ"
data_go$GOgroup[data_go$group_membrane=="1"] <- "membrane"
data_go$GOgroup[data_go$group_extra=="1"] <- "extra"

# Assign groups 4-5 based on sum column 
data_go$GOgroup[data_go$group_var_na=="2"] <- "various"
data_go$GOgroup[data_go$group_var_na=="3"] <- "various"
data_go$GOgroup[data_go$group_var_na=="0"] <- "Unknown"

# Select only necessary columns for final dataset
data_final <- data_go %>%
  select("accession_number", "GOgroup", 6:29)
```

## Summarise data

```{r}
# Change data into long format
data_long <- data_final %>%
  gather("SampleID", "percent_of_total_spectra", 3:26)

# Fix sample IDs
data_long <- data_long %>%
  separate("SampleID", c("SampleID", "Remove"), sep="_") %>%
  select(-c("Remove"))
data_long$SampleID <- toupper(data_long$SampleID)

# Remove percentage sign and change quant values to numeric
data_long <- data_long %>%
  mutate(percent_of_total_spectra = gsub("%", "", percent_of_total_spectra))
data_long$percent_of_total_spectra <- as.numeric(data_long$percent_of_total_spectra)

# Remove proteins without spectra
data_long <- data_long[-which(data_long$percent_of_total_spectra==0), ]

# Add metadata
data_meta <- left_join(metadata, data_long, by="SampleID")

# Save data for faster future plotting 
write.table(data_meta, "<PATH_TO_FILE>", row.names = FALSE)

# Summary for stats, by number of total protein clusters
cluster_summary <- data_meta %>%
  group_by(SampleID, Category, Weight, Protocol, Individual) %>%
  summarize(total_clusters=n())

# Calculate mean and sd for RUV001
cluster_summary_ruv <- cluster_summary %>%
  filter(Individual=="RUV001")
mean(cluster_summary_ruv$total_clusters)
sd(cluster_summary_ruv$total_clusters)

# Calculate mean and sd for WIG001
cluster_summary_wig <- cluster_summary %>%
  filter(Individual=="WIG001")
mean(cluster_summary_wig$total_clusters)
sd(cluster_summary_wig$total_clusters)
```

## Stats

Let's first check if there are differences in the total number of protein clusters.

```{r}
# Test without RUV001
cluster_summary_noruv <- cluster_summary %>%
  filter(!Individual=="RUV001")
cluster_signif_test_noruv <-
  pairwise.wilcox.test(cluster_summary_noruv$total_clusters,
                       cluster_summary_noruv$Category, p.adjust.method = "BH")
cluster_signif_test_noruv 
```

Now we can check if there are differences in specific GO-groups between the categories. For that, we will normalize the number of cluster per GO-group by the total number of clusters, in order to compare proportions instead of numbers.

```{r}
# Summary by GO-group, with RUV001 removed
GO_summary <- data_meta %>%
  group_by(SampleID, Category, Weight, Protocol, Individual, GOgroup) %>%
  summarize(total_clusters=n()) %>%
  filter(!Individual=="RUV001")

# Turn into wide format
GO_summary_wide <- GO_summary %>%
  spread(GOgroup, total_clusters, fill=0)

# Create column for total numer of clusters
GO_summary_wide$total <- rowSums(GO_summary_wide[ ,6:10])

# Normalize GO-groups by total number of clusters
GO_summary_wide <- GO_summary_wide %>%
  mutate_at(6:10, list(~./total))

# Change into long format
GO_norm <- GO_summary_wide %>%
  gather(GOgroup, proportion, 6:10)

# Apply statistical test to GO group
wilcox.GO <- function(x) {
  broom::tidy(pairwise.wilcox.test(x$proportion, x$Category, p.adjust.method = "BH"))
}
results_GO <- GO_norm %>%
  group_by(GOgroup) %>%
  group_map(~ wilcox.GO(.x)) 

# NOTE: Since these proportions are dependent on each other, further p-value correction is necessary. However, since nothing is significant here, no further corrections will be done in this script.
```

## Plot

Now that we have GO-groups for all identified protein clusters, we can create a plot to show them. However, we first need to transform the data into a form we can actually plot.

```{r}
# Load previously made data file
cluster_data <- read.table( "<PATH_TO_FILE>", header = TRUE)

# Order the GO groups
cluster_data$GOgroup <- factor(cluster_data$GOgroup, 
                               levels=c("intra_organ", "membrane", "extra", "various", "Unknown"))

# Set order of categories 
cluster_data$Category <- factor(cluster_data$Category, levels=c("PO10", "UP10", "PO2", "UP2"))

# Set colors per GO-category 
my_colors = c("#AA4499", "#6699CC", "#332288", "#44AA99", "#888888")
names(my_colors) <- cluster_data$GOgroup %>% unique %>% sort

# Re-order individuals from oldest to youngest
cluster_data$Individual <- factor(cluster_data$Individual, 
  levels=c("DRT001", "SMD017", "SMD046", "SMD051", "WIG001", "RUV001"))
```

Stacked barchart: 

```{r}
clusterplot <- ggplot(cluster_data, aes(Category)) +
  geom_bar(aes(fill=GOgroup), colour="black") +
  ylab("Protein clusters") +
  xlab("Category") +
  facet_grid(~Individual) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2)) +
  scale_fill_manual(values=my_colors,
              name="GO group",
              labels=c("Intracellular/organelle", "Membranes",
                       "Extracellular", "Various", "Unknown"))
```


# Top 20 protein clusters

## Import data

Input data is a cluster report exported from Scaffold for all samples.

```{r}
#Import data
raw_data <- read.delim("<PATH_TO_FILE>") %>%
  janitor::clean_names() 

# Import metadata
UP_metadata_prot <- read.delim("<PATH_TO_FILE>")
```

### Data cleanup
```{r}
# Remove decoy hits and contaminants
data <- raw_data %>%
  filter(!str_detect(protein_accession_numbers, 'rr//|REV')) %>%
  filter(!str_detect(protein_cluster, "Collagen|Keratin|Serum albumin"))

# Select only necessary columns, fix sample names and remove percentage sign from values
data <- data %>%
  select(c(ms_ms_sample_name, protein_cluster, percentage_of_total_spectra)) %>%
  separate(ms_ms_sample_name, c("SampleID", "remove"), "_") %>%
  select(-c(remove)) %>%
  mutate(percentage_of_total_spectra = gsub("%", "", percentage_of_total_spectra)) %>%
  mutate(percentage_of_total_spectra = as.numeric(percentage_of_total_spectra))
```

### Find 20 most abundant protein clusters

RUV001 will not be considered in seleting most abundant protein clusters, since it does not exhibit the same calculus profile as the other samples.

```{r}
# Turn data into wide format
data_wide <- data %>%
  spread(SampleID, percentage_of_total_spectra, fill=0)

# Remove RUV samples 
data_wide_noruv <- data_wide %>%
  select(-c(RUV001A04, RUV001A07, RUV001A15, RUV001A16))

# Remove proteins only found in RUV001 samples
data_wide_noruv <- data_wide_noruv[rowSums(data_wide_noruv[2:21]) !=0, ]

# Calculate average percentage
data_wide_noruv$ave_perc <- rowMeans(data_wide_noruv[2:21])

# Order by average percentage
data_wide_noruv <- data_wide_noruv[order(-data_wide_noruv$ave_perc),]

# Choose 20 most abundant protein clusters
top20_noruv <- data_wide_noruv[1:20, ]
```

## Statistics

Significant differences in abundance will be tested by doind a pairwise Wilcoxon rank sum test with Benjamini-Hochberg correction for each of the top 20 species.

```{r}
# Select only species column and average percentage 
top20 <- top20_noruv %>%
  select(c(protein_cluster, ave_perc))

# Subset data for only top 20 species
top20_data <- left_join(top20, data_wide)

# Change data into long format and remove average
top20_data_long <- top20_data %>%
  gather(SampleID, percentage_spectra, 3:26) 

# Add metadata
top20_long_meta <- left_join(top20_data_long, UP_metadata_prot)

# Fix protein cluster ID's
top20_long_meta <- top20_long_meta %>%
  mutate(protein_cluster = gsub("Cluster of ", "", protein_cluster)) %>%
  separate(protein_cluster, c("protein_cluster", "remove"), "GN") %>%
  select(-c(remove)) %>%
  separate(protein_cluster, c("protein_cluster", "taxa"), "OS=")

# Save for faster plotting in the future
write.table(top20_long_meta,  "<PATH_TO_FILE>",  row.names = FALSE)

# Remove RUV001
top20_long_meta_noruv <- top20_long_meta %>%
  filter(!Individual=="RUV001")

# Apply statistical test to each protein cluster 

### NOTE: This is still done one protein cluster at a time, I need to ask someone R-proficient how to make it into a loop ###

pc <- top20_long_meta_noruv %>%
  filter(protein_cluster=="<PROTEIN>")
pc_test <- tidy(pairwise.wilcox.test(pc$percentage_spectra, pc$Category, p.adjust.method = "BH"))

# Add sample to list of all protein clusters
pc_test$protein_cluster <- "<PROTEIN>"
pc_results <- rbind(pc_test, pc_results)

pc_results <- pc_test #Only needs to be done first time 

# Save results
write.table(pc_results, " "<PATH_TO_FILE>"")

### NOTE: You will need to adjust for each protein comparison too, since they are not independent. ###

# Extract significant results
sign_top20 <- pc_results %>%
  filter(p.value < 0.05)

# BH correction for 20 dependent proteins 
p.adjust(sign_top20$p.value, method = "BH", n = 20)
```

## Plotting

Data will be presented as a heatmap of protein cluster abundances across all samples.

```{r}
# Read in data
top20_long_meta <- read.table("<PATH_TO_FILE>", header = TRUE)

# Set order of categories 
top20_long_meta$Category <- factor(top20_long_meta$Category, levels=c("PO10", "UP10", "PO2", "UP2"))

# Re-order individuals from oldest to youngest
top20_long_meta$Individual <- factor(top20_long_meta$Individual, 
  levels=c("DRT001", "SMD017", "SMD046", "SMD051", "WIG001", "RUV001"))
```

Create plot:

```{r}
top20plot <- ggplot(top20_long_meta, 
      aes(x=Category, y=reorder(protein_cluster, percentage_spectra), fill = percentage_spectra)) + 
  geom_tile() + 
  coord_equal() +
  scale_fill_gradient(low = "white", high = "#332288", 
                      name="Percentage of total spectra") +
  theme_minimal(base_size = 10) + 
  theme(axis.text.x = element_text(angle = 90, hjust=0.95, vjust=0.2)) +
  theme(axis.text=element_text(size=7)) +
  ylab("") +
  facet_grid(~ Individual) 
```
Proteins will be coloured by origin (human/bacterial/eukaryotic) manually.


# Principal component analysis

## Preparation

The input data is a "cluster report" exported from Scaffold, filtered for 5% protein FDR  and 1% peptide FDR.

```{r import, warning=FALSE}
# Import data
raw_data <- read.delim("<PATH_TO_FILE>") %>%
  janitor::clean_names() 

# Import metadata
UP_metadata_prot <- read.delim("<PATH_TO_FILE>")
```

## Data preparation

```{r cleanup, results='hide', message=FALSE, warning=FALSE}
# Remove any decoy hits
data <- raw_data %>%
  filter(!str_detect(protein_accession_numbers, 'rr\\|REV')) 

# Remove collagens and keratins
data <- data %>%
  filter(!str_detect(protein_cluster, "Collagen|Keratin|Serum albumin"))

# Remove unnecessary parts from sample name
data <- data %>%
  separate(ms_ms_sample_name, c("SampleID", "remove"), "_") %>%
  dplyr::select(-c(remove))
```

## Filtering 

For the PCA we will filter for presence of the cluster in at least two samples, to remove spurious (possibly false) hits, and use percentage of total spectra directly as quatitative value. Pseudocount zero replacement is done, and the data is then log2-transformed. This is following Jersie-christensen 2018 with some modifications. Two different PCA's are done, one with and one without RUV001.

### Prepare data
```{r norm4, results='hide', message=FALSE, warning=FALSE}
# Select only the columns we need for further analysis
data_pca <- data %>% 
  dplyr::select(c("SampleID", "protein_cluster",
                  "percentage_of_total_spectra"))

# Remove percentage sign
data_pca$percentage_of_total_spectra <- as.numeric(gsub("%", "",
                            data_pca$percentage_of_total_spectra))

# Transform data to wide format
data_pca_wide <- spread(data_pca, SampleID, 
                        percentage_of_total_spectra, fill = NA)

# Fix protein cluster names
data_pca_wide <- data_pca_wide %>%
  mutate(protein_cluster = gsub("Cluster of ", "", protein_cluster)) %>%
  separate(protein_cluster, c("protein_cluster", "remove"), "GN") %>%
  dplyr::select(-c(remove))

# Create dataset without RUV001, remove protein that were only present in RUV001
data_pca_wide_noruv <- data_pca_wide %>%
  dplyr::select(-starts_with("RUV"))
data_pca_wide_noruv <- data_pca_wide_noruv[rowSums(is.na(data_pca_wide_noruv[2:21]))!=20,]

# Only keep proteins present in at least two samples
data_pca_wide <- data_pca_wide[rowSums(is.na(data_pca_wide[2:25]))<=22,]
data_pca_wide_noruv <- data_pca_wide_noruv[rowSums(is.na
                                  (data_pca_wide_noruv[2:21]))<=18,]

# Change NA values into 0
data_pca_wide[is.na(data_pca_wide)] <- 0
data_pca_wide_noruv[is.na(data_pca_wide_noruv)] <- 0

# Pseudocount zero replacement to make log-transformation possible
data_pca_wide[,2:ncol(data_pca_wide)] <-
  data_pca_wide[,2:ncol(data_pca_wide)]+1
data_pca_wide_noruv[,2:ncol(data_pca_wide_noruv)] <- 
  data_pca_wide_noruv[,2:ncol(data_pca_wide_noruv)]+1

# log-transform spectral values
data_pca_wide[, 2:25] <- log2(data_pca_wide[2:25])
data_pca_wide_noruv[, 2:21] <- log2(data_pca_wide_noruv[2:21])
```

A total of 102 protein clusters survive filtering both with and without RUV001. 

## PCA
```{r PCA4, results='hide', message=FALSE, warning=FALSE}
# Transpose dataframe
data_pca_wide_tr <- data_pca_wide %>% 
  gather(SampleID, value, 2:ncol(.)) %>%
  spread(protein_cluster, value)
data_pca_wide_tr_noruv <- data_pca_wide_noruv %>% 
  gather(SampleID, value, 2:ncol(.)) %>%
  spread(protein_cluster, value) 

# Create PCA object
pca.all <- prcomp(data_pca_wide_tr[, 2:102], scale=FALSE)
pca.noruv <- prcomp(data_pca_wide_tr_noruv[, 2:102], scale=FALSE)
```

## Plotting

Collect values from PCA objects.
```{r}
# Extract values
pca_all_out <- as_tibble(pca.all$x)
pca_all_out$SampleID <- data_pca_wide_tr$SampleID

pca_noruv_out <- as_tibble(pca.noruv$x)
pca_noruv_out$SampleID <- data_pca_wide_tr_noruv$SampleID

# Add metadata 
pca_all_meta <- left_join(UP_metadata_prot, pca_all_out)
pca_noruv_meta <- left_join(pca_noruv_out, UP_metadata_prot)
```

To see what is driving the variation in the samples, we can take a closer look at the loadings of the PCA object.

```{r}
# Get variables with biggest impact for PC1 and PC2
pca.all$rotation[,1] %>% sort() %>% head(2)
pca.all$rotation[,1] %>% sort(decreasing = TRUE) %>% head(2)
pca.all$rotation[,2] %>% sort() %>% head(2)
pca.all$rotation[,2] %>% sort(decreasing = TRUE) %>% head(2)

# And same for PCA without RUV001
pca.noruv$rotation[,1] %>% sort() %>% head(2)
pca.noruv$rotation[,1] %>% sort(decreasing = TRUE) %>% head(2)
pca.noruv$rotation[,2] %>% sort() %>% head(2)
pca.noruv$rotation[,2] %>% sort(decreasing = TRUE) %>% head(2)

# Get rownames separately for each protein
which(rownames(pca.noruv$rotation) %in% c("<PROTEIN>", "<PROTEIN>"))

# Create matrix of x (PC1) coordinates and y (PC2) (duplicates removed)
l.x <- cbind(pca.all$rotation[,1][c(17, 55, 45, 67, 48, 92,78)]) #17 twice
l.y <- cbind(pca.all$rotation[,2][c(17, 55, 45, 67, 48, 92, 78)]) #17 twice

# Create matrix of coordinates without RUV001 (duplicates removed)
l.x.2 <- cbind(pca.noruv$rotation[,1][c(17, 59, 45, 67, 48, 57, 78)]) #17 twice
l.y.2 <- cbind(pca.noruv$rotation[,2][c(17, 59, 45, 67, 48, 57, 78)]) #17 twice
```

Set aesthetics.
```{r}
# Calculate percent explained per PC
percentage_all <- round(pca.all$sdev / sum(pca.all$sdev) * 100, 2)
percentage_all <- paste(colnames(pca_all_out), "(", paste(
  as.character(percentage_all), "%", ")", sep="") )

percentage_noruv <- round(pca.noruv$sdev / sum(pca.noruv$sdev) * 100, 2)
percentage_noruv <- paste(colnames(pca_noruv_out), "(", paste(
  as.character(percentage_noruv), "%", ")", sep="") )

# Set colours for individuals
my_colors = c("#999933", "#888888", "#6699CC", "#332288", "#44AA99", "#AA4499")
names(my_colors) = pca_all_meta$Individual %>% unique %>% sort

# Set shapes for categories
my_shapes = c(19, 1, 17, 2)
names(my_shapes) = pca_all_meta$Category %>% unique %>% sort
```

And finally create plots!
```{r}
pca_all <- ggplot(data = pca_all_meta, 
       aes(x = PC1, y = PC2, colour=Individual, shape=Category)) +
  geom_point(size=4, stroke=3) +
  scale_shape_manual(values=my_shapes) +
  scale_color_manual(values=my_colors) +
  xlab(percentage_all[1]) + 
  ylab(percentage_all[2]) +
  guides(fill=guide_legend(override.aes=list(shape=21, size=5))) +
  theme_minimal(base_size = 12) +
  annotate("segment", x=0, xend=l.x, y=0, yend=l.y, colour="black", 
           size=1, arrow=arrow()) + 
  annotate("text", x=l.x, y=l.y, label=rownames(l.x), color="red")

pca_noruv <- ggplot(data = pca_noruv_meta, 
       aes(x = PC1, y = PC2, colour=Individual, shape=Category)) +
  geom_point(size=4, stroke=3) +
  scale_shape_manual(values=my_shapes) +
  scale_color_manual(values=my_colors) +
  xlab(percentage_noruv[1]) + 
  ylab(percentage_noruv[2]) +
  guides(fill=guide_legend(override.aes=list(shape=21, size=5))) +
  theme_minimal(base_size = 10) +
  annotate("segment", x=0, xend=l.x.2, y=0, yend=l.y.2, colour="black", 
           size=1, arrow=arrow()) + 
  annotate("text", x=l.x.2, y=l.y.2, label=rownames(l.x.2), color="firebrick")
```

## PERMANOVA

In order to see if the groupings are significantly different from each other, we will perform a PERMANOVA using the R package 'vegan'. Input data is sequence counts normalized in the same way as in the PCA above.

```{r}
# Remove RUV001 from metadata file
meta_noruv <- UP_metadata_prot %>%
  filter(!Individual=="RUV001")
```

And then we run the test.
```{r}
# Basic permanova with euclidean distance as metric
permanova <- adonis(data_pca_wide_tr_noruv[, 2:103] ~ Individual + Weight + Protocol, 
                    data=meta_noruv,
                    permutations=999,
                    method="euclidean")
permanova
```

We also need to check that the variance homogeneity assumptions hold.

```{r}
dist <- vegdist(data_pca_wide_tr_noruv[, 2:103])
anova(betadisper(dist, meta_noruv$Protocol))
```

Looks good!

# Combine plots

These three plots will be in the same figure in the paper. The PCA with RUV001 will be placed in the supplement.

```{r}
fig8 <- plot_grid(clusterplot, 
                  plot_grid(top20plot, pca_noruv, labels = c('B', 'C'), 
                            label_size = 12, ncol = 2),
                  labels = c('A', ''), label_size = 16, nrow = 2)
```

