---
title: "Protein characteristics"
output: html_notebook
author: "Zandra Fagernaes"
---

This is a walkthrough of all the protein characteristics analyses for the paper "A unified protocol for simultaneous extraction of DNA and proteins from archaeological dental calculus" (Fagernaes et al. 2020). The data is summarized in figure 7 and supplementary material. Individual RUV001 in not included in any statistics, as stated in the paper.

```{r echo = FALSE}
library(janitor)
library(fuzzyjoin)
library(readxl)
library(reshape2)
library(gtools)
library(ggsignif)
library(ggpubr)
library(Peptides)
library(tidyverse)
```

## Hydrophobicity

Here, we will investigate the hydrophobicity of the proteins identified through LC-MS/MS. The web application GRAVY (http://www.gravy-calculator.de) was used to calculate hydrophobicity (see details in paper).

### Import data

```{r}
# Import protein report
protein_report <- read.delim("<PATH_TO_FILE>") %>%
  janitor::clean_names() 

# Results from GRAVY
gravy <- read_excel("<PATH_TO_FILE>") %>%
  janitor::clean_names() 

# Metadata
UP_metadata_prot <- read.delim("<PATH_TO_FILE>")
```

Clean GRAVY file

```{r}
# Remove reverse hits and keep only Uniprot identifier
gravy <- gravy %>%
  filter(!str_detect(header, 'rr//|REV')) %>%
  separate(header, c("accession", "remove"), " ") %>%
  separate(accession, c("r1", "r2", "protein_accession_numbers"), "[|]") %>%
  dplyr::select(-c(remove, r1, r2)) 
```

Clean protein report file

```{r}
# Select only necessary columns
data <- protein_report %>%
  dplyr::select(c(ms_ms_sample_name, protein_accession_numbers, 
           percentage_of_total_spectra, protein_name))

# Fix sample IDs
data <- data %>%
  separate(ms_ms_sample_name, c("SampleID", "remove"), " ") %>%
  separate(SampleID, c("SampleID", "remove2"), "_") %>%
  dplyr::select(-c(remove, remove2)) 

# Remove reverse hits and contaminants
data <- data %>%
  filter(!str_detect(protein_accession_numbers, 'rr//|REV')) %>%
  filter(!str_detect(protein_name, "Collagen|Keratin|Serum albumin"))
```

### Assign hydropathy value to identified proteins

Since the protein report contains multiple accessions for each protein, we can't just join the datasets based on accession. The R package fuzzyjoin is thereby used to use regular expressions to assign a hydropathy value to each protein in the protein report. 

In cases where several protein accessions are given for a protein (i.e. when several proteins are matched), the mean hydropathy of these proteins are used.

```{r}
# Combine datasets
data_gravy <- regex_left_join(data, gravy, by="protein_accession_numbers")

# Average over proteins when several accessions are given
data_gravy <- data_gravy %>%
  group_by(SampleID, protein_accession_numbers.x, percentage_of_total_spectra) %>%
  summarize(mean_gravy = mean(gravy))

# Add metadata
data_gravy_meta <- left_join(data_gravy, UP_metadata_prot)

# Calculate mean per individual and category
data_gravy_summary <- data_gravy_meta %>%
  group_by(SampleID, Category, Weight, Protocol, Individual) %>%
  summarize(mean_hydro=mean(mean_gravy, na.rm=TRUE))
```

### Stats

Significances of changes in mean hydrophobicity between the protocols/starting weights are calculated with a pairwise Wilcoxon test, and corrected with the Benjamini-Hochberg method for multiple testing.

```{r}
# Remove RUV001 and test
data_gravy_summary_noruv <- data_gravy_summary %>%
  filter(!Individual=="RUV001")
hydro_signif_test_noruv <- pairwise.wilcox.test(data_gravy_summary_noruv$mean_hydro, data_gravy_summary_noruv$Category, p.adjust.method = "BH")
hydro_signif_test_noruv 
```

### Plot boxes

Set aesthetics

```{r}
# Set colors per individual
my_colors = c("#999933", "#888888", "#6699CC", "#332288", "#44AA99", "#AA4499")
names(my_colors) = data_gravy_summary$Individual %>% unique %>% sort

# Function for getting same number of decimals in plots
formatter <- function(...){
  function(x) format(round(x, 1), ...)
}
```

Plot!

```{r}
hydro_plot <-ggplot(data_gravy_summary, aes(x=Category, y=mean_hydro)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(position=position_jitterdodge(dodge.width=0.3),
              aes(color=Individual), shape= 19, size=4) +
  scale_color_manual(values=my_colors) +
  scale_x_discrete(limits=c("PO10", "UP10", "PO2", "UP2")) +
  ylab("Hydropathy index") +
  theme_minimal(base_size = 12)
```


## Amino acid composition

In order to make sure that the protocol is not biasing against some type of amino acids, the amino acid composition of the identified peptides was calculated. This was done on a peptide report exported through Scaffold.

```{r}
# Load and clean data
peptide_report <- read.delim("<PATH_TO_FILE>") %>%
  janitor::clean_names() %>%
  select("biological_sample_name", "ms_ms_sample_name", "protein_name", "protein_accession_numbers", "percentage_of_total_spectra", "peptide_sequence") %>%
  separate(ms_ms_sample_name, c("SampleID", "remove"), " ") %>%
  select(-c(remove)) %>%
  separate(SampleID, c("SampleID", "remove"), "_") %>%
  select(-c(remove)) 

# Remove reverse hits and contaminants 
peptide_report <- peptide_report %>%
  filter(!str_detect(protein_accession_numbers, 'rr//|REV')) %>%
  filter(!str_detect(protein_name, "Collagen|Keratin|Serum albumin"))

# Load metadata
UP_metadata_prot <- read.delim("<PATH_TO_FILE>")
```

Now we need to create separate datasets for each sample, deduplicate the peptides and concatenate them into one long sequence. We can then analyze this "super-peptide" with the Peptides package (Osorio et al. 2015).

**NOTE: This is still done one sample at a time, I need to ask someone R-proficient how to make it into a loop**

```{r}
# Create data set per sample
sample <- peptide_report %>%
  filter(SampleID=="<SAMPLEID>") %>%
  distinct(peptide_sequence, .keep_all = TRUE)

# Concatenate all peptides into one
super_pep=NULL
for (i in sample$peptide_sequence) super_pep<-paste0(super_pep,i)

# Calculate amino acid composition of super-peptide
sample_aa <- as.data.frame(aaComp(super_pep)) %>%
  tibble::rownames_to_column("aa_type")

# Add sample to list of all amino acid compositions
sample_aa$SampleID <- "<SAMPLEID>"
aa_comp <- rbind(aa_comp, sample_aa)

aa_comp <- sample_aa #Only needs to be done first time 
```

Now save the file so it doesn't need to be done every time.

```{r}
write.table(aa_comp, "<PATH_TO_FILE>")
```

And now let's check if there are any statistical differences in the proportion of each amino acid type with a pairwise Wilcoxon test, corrected with the Benjamini-Hochberg method for multiple testing.

```{r}
# Load previously created file
UP_peptide_aa_comp <- read.csv("<PATH_TO_FILE>", sep="")

# Add metadata
aacomp_meta <- left_join(UP_peptide_aa_comp, UP_metadata_prot)

# Remove RUV001
aacomp_meta_noruv <- aacomp_meta %>%
  filter(!Individual=="RUV001")

### NOTE: This is still done one amino acid type at a time, I need to ask someone R-proficient how to make it into a loop ###

# Apply statistical test to each amino acid type
aa <- aacomp_meta_noruv %>%
  filter(aa_type=="Basic")
aa_test <- tidy(pairwise.wilcox.test(aa$Mole., aa$Category, p.adjust.method = "BH"))

# Add sample to list of all amino acid compositions
aa_test$aa_type <- "Basic"
aa_results <- rbind(aa_test, aa_results)

aa_results <- aa_test #Only needs to be done first time 
```

Only the basic amino acids will be shown in the plot, since they have the clearest pattern. Acidic amino acids will be shown in supplementary figure.

```{r}
### Basic

# Subset to only basic amino acids
basic_aa <- aacomp_meta %>%
  filter(aa_type=="Basic")

# Extract only significant comparisons
aa_results_basic <- aa_results %>%
  filter(aa_type == "Basic")
significant_aa_basic <- aa_results_basic$p.value %>% melt() %>% filter(value < 0.05)

# Add asterisks to represent significance
significant_aa_basic$stars <- stars.pval(significant_aa_basic$value)

### Acidic

# Subset to only acidic amino acids
acidic_aa <- aacomp_meta %>%
  filter(aa_type=="Acidic")

# Extract only significant comparisons
aa_results_acidic <- aa_results %>%
  filter(aa_type == "Acidic")
significant_aa_acidic <- aa_results_acidic$p.value %>% melt() %>% filter(value < 0.05)

# Add asterisks to represent significance
significant_aa_acidic$stars <- stars.pval(significant_aa_acidic$value)
```

Now this can be added to a plot.

```{r}
# Set colors per individual
my_colors = c("#DDCC77", "#888888", "#6699CC", "#332288", "#44AA99", "#AA4499")
names(my_colors) <- basic_aa$Individual %>% unique %>% sort

# Function for getting same number of decimals in plots
formatter <- function(...){
  function(x) format(round(x, 1), ...)
}
```

Plot!
```{r}
basic_plot <- ggplot(basic_aa, aes(x=Category, y=Mole.)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitterdodge(dodge.width=0.3),
              aes(color=Individual), shape=19, size=4) +
  scale_color_manual(values=my_colors) +
  geom_signif(comparisons = list(c("UP10", "PO10"), c("UP2", "PO10"),
                                 c("UP10", "PO2"), c("UP2", "PO2")),
              annotations = significant_aa_basic$stars, 
              y_position = c(15, 15.6, 15.2, 15.4)) +
  scale_x_discrete(limits=c("PO10", "UP10", "PO2", "UP2")) +
    scale_y_continuous(labels = formatter(nsmall = 1)) +
  ylab("Basic amino acids (%)") +
  theme_minimal(base_size = 12)
```

Plot!
```{r}
acidic_plot <- ggplot(acidic_aa, aes(x=Category, y=Mole.)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitterdodge(dodge.width=0.3),
              aes(color=Individual), shape=19, size=4) +
  scale_color_manual(values=my_colors) +
  geom_signif(comparisons = list(c("UP2", "PO10")),
              annotations = significant_aa_acidic$stars, 
              y_position = 14.5) +
  scale_x_discrete(limits=c("PO10", "UP10", "PO2", "UP2")) +
  scale_y_continuous(labels = formatter(nsmall = 1)) +
  ylab("Acidic amino acids (%)") +
  theme_minimal(base_size = 12)
```


## Molecular weight

Input data is the "Protein report" from Scaffold. This will be a supplementary figure.

### Load data

```{r}
# Protein report
protein_report <- read.delim("<PATH_TO_FILE>") %>%
  janitor::clean_names() 

# Metadata
UP_metadata_prot <- read.delim("<PATH_TO_FILE>")
```

### Data cleanup

```{r}
# Fix sample IDs
data <- protein_report %>%
  separate(ms_ms_sample_name, c("Sample", "remove"), " ") %>%
  separate(Sample, c("SampleID", "remove2"), "_") %>%
  dplyr::select(-c(remove, remove2)) 

# Remove reverse hits and contaminants 
data <- data %>%
  filter(!str_detect(protein_accession_numbers, 'rr//|REV')) %>%
  filter(!str_detect(protein_name, "Collagen|Keratin|Serum albumin"))

# Change weight column into numeric and make it kDa
data$protein_molecular_weight_da <- as.numeric(data$protein_molecular_weight_da)
data$protein_molecular_weight_kda <- data$protein_molecular_weight_da/1000
```

### Calculate means

```{r}
# Select only necessary columns
data_weight <- data %>%
  dplyr::select(SampleID, protein_name,
                protein_molecular_weight_kda)

# Add metadata
data_weight_meta <- left_join(data_weight, UP_metadata_prot)

# Calculate mean per individual and category
data_weight_summary <- data_weight_meta %>%
  group_by(SampleID, Category, Weight, Protocol, Individual) %>%
  summarize(mean_weight=mean(protein_molecular_weight_kda, na.rm=TRUE))

# Calculate mean and sd for RUV001
data_weight_summary_ruv <- data_weight_summary %>%
  filter(Individual=="RUV001")
mean(data_weight_summary_ruv$mean_weight)
sd(data_weight_summary_ruv$mean_weight)

# Calculate mean and sd for other individuals
data_weight_summary_noruv <- data_weight_summary %>%
  filter(!Individual=="RUV001")
mean(data_weight_summary_noruv$mean_weight)
sd(data_weight_summary_noruv$mean_weight)
```

### Stats

Significances of changes in mean molecular weight between the protocols/starting weights are calculated with a pairwise Wilcoxon test, and corrected with the Benjamini-Hochberg method for multiple testing.

```{r}
# Without RUV001
data_weight_summary_noruv <- data_weight_summary %>%
  filter(!Individual=="RUV001")
weight_noruv_signif_test <- pairwise.wilcox.test(data_weight_summary_noruv$mean_weight, data_weight_summary_noruv$Category, p.adjust.method = "BH")
weight_noruv_signif_test 
```

### Box plot

Set aesthetics
```{r}
# Set colors per individual
my_colors = c("#DDCC77", "#888888", "#6699CC", "#332288", "#44AA99", "#AA4499")
names(my_colors) <- data_weight_meta$Individual %>% unique %>% sort
```

Plot!
```{r}
weight_plot <- ggplot(data_weight_summary, aes(x=Category, y=mean_weight)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitterdodge(dodge.width=0.3),
              aes(color=Individual), shape=19, size=4) +
  scale_color_manual(values=my_colors) +
  scale_x_discrete(limits=c("PO10", "UP10", "PO2", "UP2")) +
  ylab("Molecular weight (kDa)") +
  theme_minimal(base_size = 12)
```


## Combine plots

Since these analyses will be in the same section of the paper, the plots will be combined into one figure. 

```{r echo = FALSE}
prot_hydro_fig <- ggarrange(hydro_plot, basic_plot,
                           labels = c("A", "B"),
                           ncol = 1, nrow = 2, 
                           common.legend = TRUE, 
                           legend = "right")

```
